Querying Integrated Scientific Observation and Measurement Data

### 2010-05-07 docs/algdraft/DataIntegration/DataIntegration.tex
Refine the definition and add introduction. Is refining the materialization part.

\section{Introduction}
In this work, we study the query processing over integrated ecological
data through OBOE model\cite{DBLP:conf/er/BowersMS08}, a conceptual model used to interpret
observation and measurement data.

\subsection{Background}
Introduce OBOE, annotation
Give an example of them.

@@

@@

In many scientific domains (e.g., ecology, hydrology, earth science, geology), people collect observation and
measurement data. Such data record the observed value of some real
world entity at some specific place. E.g., ecologist studying the
growth pattern and treatments often need to record the tree height,
The collected data reflect the fact the tree height of a specific tree is 1.0in on
May 1, 2009 and 1.5in on May 1, 2010. ({\bf check whether this is
  realistic or not.})

In scientific domains where people collect observation and measurement data, there are
several commonly used and widely recognized canonical
concepts(\cite{oboe, om}). 
In this paper, we refer these concepts as OM concetps. 
The canonical concepts include {\em observation}, {\em measurement},
{\em characteristic}, {\em standard}, {\em protoca} or ({\em procedure}). 
For example, {\bf add an example to illustrate these concepts. }

When a user poses a query, it is unrealistic for a him/her to know the 
underlying data structure of the data.  
But they all know the well recognized OM conceps as discussed above. 
So, naturally, when searching such scientific data, people are more interested in finding datasets
related to such concepts, e.g., specific observation and measurement.

@@

When a scientist contributes dataset into the data repository, generally
each dataset is converted to a data table\cite{tdar} or treated as a separate
object entity\cite{knb}. 
So, the database contains metadata about each dataset, the data table
definition (e.g., attribute/column name, attribute type, etc). 
%\from{HP}{Another way to organize the data is to store them in one big
%data table with three columns data\_file, attribute\_name,
%attribute\_value. The problem with this organization is that the
%attribute\_value are different types. } 
To formalize the scenario, we use $D$ to denote the set of data
tables. Each data table contains metadata about the attribute
definition.  Sometimes, we also use {\em column} to denote the attribute. 
Let $d$ be one data table, 
$d[attr]$ or $d[idx]$ represents the set of values for the attribute
$attr$ or for the $idx$-th attribute. 
Besides the dataset, in the data store, we also keep the annotation
information. 
%Implementation detail (no need to put it in the paper): 
%When we put the data to data tables, we add a unique record id for
%each row in the data. This is used to get the unique record id.

\subsection{Annotation}\label{sec:annotation}.
In the Ecology domain, people are using annotations to connect
datasets to the OBOE model. 
{\bf TODO: Put a screen-dump of annotation.}
When some annotations are done over data tables, the system keeps the
information $\langle AnnotId, DTableId\rangle$.

Map annotation information 
We use $A$ to denote the annotation of one dataset. 
More specifically,
To describe an observation, we need to denote on which entity (object in the
real world) the observation is made.
Very often, more than one observation can be made on one entity.   
\begin{itemize}
\item A.ObsType = $\{\langle \underline{AnnotId,ObsTypeId}, EntTypeId,
  isDistinct \rangle\}$ 
\item A.MeasType = $\{\langle \underline{AnnotId, MeasTypeId}, ObsTypeId,
  CharType, StdType, ProtType, Precision, isKey \rangle\}$ 
\item A.ContextType = $\{\langle \underline{AnnotId, ObsTypeId, ContextObsTypeId, RelType}, isIdentify\rangle\}$
\item A.Map = $\{\langle \underline{AnnotId,MeasTypeId, ResAttribute, Cond}, Val\rangle\}$
\end{itemize}

``AnnotId'' denotes the resource (e.g., data table) the annotation is on.

{\bf TODO: Describe the constraints in annotation}

\subsection{Observation and measurement (OM) query}

{\bf TODO: Have a running example from here.}
Observation and measurement (OM) query

\begin{definition}[Basic OM query]\label{def:basic_omq}
A basic observation and measurement query is defined as $Q$::={\tt
  concept:cond}. 
Here, {\tt concept} is the main term in an observation model. 
In using the OBOE model, it can be observation, measurement,
characteristic or standard. 
{\tt cond} is in the form of $f([distinct]~attr)~op~value$ where $op$ is
basic operator from $\{=, \neq, >, <\}$ and $f$ is an
aggregation function from $\{sum, avg, count, min, max\}$. 
\end{definition}

\begin{definition}\label{def:qresult}
The result of a query $Q$ is a set of data objects (e.g., data tables) 
$\{d|d\in D \wedge d~satisfies~{\tt cond}\}$. 
%$\{d|d\in D \wedge (value~op~dv = true, dv\in d[attr])\}$. 
For each of such result $d$, we use $d~s.t.~Q$ to denote that
data object $d$ satisfies the query. 
\end{definition}

Definition \ref{def:basic_omq} can formulate queries on the
one specific observation and measurement. 
However, it fails to catch the observation context relationship. 
E.g., {\bf TODO: }
So, we generalize the basic OM query to contextualized OM query as
follows.

@@

Given the dataset in Figure \ref{}. 
People may ask the following queries.

\begin{itemize}
\item $Q_1$: Give me the datasets that contain species ``Picea rubens'' observations.
%$Q_2$: Give me the datasets that have measurements on ``area''  characteristics. 
\item $Q_2$: Give me the datasets that contain species ``Picea rubens''
  observations in ``California''. 
\item $Q_3$: Give me the dataset that contains at least five distinct ``Picea rubens'' observations.
\item $Q_4$: Give me the datasets that have trees with average ``height''
  than 20in in year 2009. 
% Shawn: an example on summarize one measure with respect to another.
% This could take two forms: one where the context forms the groups
% (like by year), and another where the two measurements are within the
% same context (like finding correlations).
\end{itemize}

They can be formalized as the following queries.

\begin{example}
The given example queries can be represented in the formal query
format. 
\begin{itemize}
\item $Observation: \langle Species='Picea~rubens'\rangle$
\item $Observation_1: \langle Species='Picea~rubens'\rangle$\\
  $IN(Observation_1, Observation_2$) \\
$Observation_2: \langle Meas_{unknown}=`California'\rangle$
\item $Observation: \langle Species='Picea~rubens' $\\
$\wedge
  count(distinct~Species)\geq 5\rangle$
\item $??$
\end{itemize}

\end{example}

\section{Naive method}\label{sec:naive}
A very direct way to help searching would be to index the data with
the data with all the values, then search over these values. 
This may get wrong results.

{\bf TODO: An example to show that the naive method give wrong answers.}
{\bf Experiement: show the error rate (e.g., precision, recall, (or
  directly $F_1$) of this method.}

@@

As we analyzed in Section \ref{sec:relatedwork}, there are two extrems
in querying integrated data. 
One is to rewrite the original query to a series of new queries over
the data; the other is to materialize the data to a consistent data
model and then answer the query over the materialized database. 
In Section \ref{sec:matdb} and \ref{sec:queryrewrite}, we introduce
our strategies in answering the queries.

%\subsection{Framework and operators}

@@

As described in Section \ref{sec:dataquery}, 
a user's query generally does not contain the underlying data
structure because the user cannot get such information. 
On the contrary, they formulate their query with the well recognized
OM concept terminologies such as observation
measurement, characteristic, etc. 
To answer such query from the original data model (i.e., dataset
metadata and data tables), we need to {\em rewrite} the quey over the
real data table. 
In what follows, we first sketch the re-writing process. Then, we
detailed the procedure in using the data model and
structures. Finally, we include the process for the more complicated
cases (e.g., with distinct, aggregate).

@@

The first step is to map the OM query to the real data structure. 
We can utilize the annotation structure $A.Map$ which keeps the correspondences between
measurement type and table attributes. 
So, when the given concept is {\em measurement}, we can directly get
its related attribute names and annotation id $AnnotId$. 
In case that the given OM concept is a non-measurement concept, we can use
other Annotation structures $A.*$ to figure out the measurementt types. 
E.g., if the OM concept is Characteristic ``Species'', we can use
$A.MeasType$ to get the measurement type id.

As mentioned in Section \ref{sec:annotation}, when some annotations are done over data tables, the system keeps the
information $\langle AnnotId, DTableId\rangle$. 
With the $AnnotId$, we can ge the data table information from the
metadata  $\langle AnnotId, DTableId\rangle$. 
Then, we get the relevant data table information and the needed attribute
information.

@@

The basic idea is to materialize the data into some centralized
concept tables with instances of entities, observations, measurements
and the contexts between observations. 
We call such concept tables core OBOE tables, and denote them as
$OBOE.*$. 
In partibular, we have the following concept instance tables. 
\begin{itemize}
\item $OBOE.EntityIns=\{\langle \underline{EntId}, EntType\rangle\}$ for all the entity instances;
\item $OBOE.ObsIns=\{\langle \underline{ObsId}, ObsType, EntId \rangle\}$ keeps all the observation instances;
\item $OBOE.MeasIns=\{\langle \underline{MeasId}, ObsId, MeasType, Val\rangle\}$ for all the measurement instances;
\item $OBOE.ContextIns =\{\langle \underline{ObsId, ConextObsId, ContextType}\rangle\}$ for all the context instances;
\end{itemize}

@@

\begin{example} \label{eg:materialize_db_q1}
Example \ref{eg:query_reqerite_q1} shows how to deal with $Q_1$ using
the query rewriting method. 
Using the  materialized database, the first step is the same.
After we know the data file, then we search the table {\em measurement\_instance} with the {\em mvalue=``Picea rubens''}. 
\end{example}

\begin{example}
We use $SQ_1$ as an example.

The first step is to find the datasets that contain the metadata. 
The second step is to perform a query over the materialized database. 
%select mlabel, mvalue, count(*) as cnt from measurement_instance
%group by mvalue,mlabel having count(*)>=2;

Using the query rewriting, 
First, we find the data tables that contain the needed concepts. 
In each table, we need to figure out the distinct observations. 
From $m_i$ $obstype_i$, all the other measurements that are of
$obstype_i$. 
Group by the key measurements of $obstype_i$ if distinct yes.
If this observation type does not have any context, then the key
measurements are all the measurements directly defined under this
observatioin type. Otherwise, i.e., this observation type has context,
then the key measurements are all the measurements in the context
chain. 
\end{example}

\begin{example}
KQ3: Give me the datasets that contain species ``Picea rubens''
  observations in ``California''.

Species = ``Picea rubens'' and observation has context {\em in}
another observation which has measurement of value ``California''. 
\end{example}

Using materializedDB, we can answer the query this way
First, Species --> mlabel, observation type

select mtype, olabel from measurement\_type  \\
where chracteristic = 'Species';  \\

For the olabel found, \\
select context\_olabel\\
from context\_relationship\\
where olabel = 'olabel';

context\_olabel\_set is a set containing all the observation type
labels in the context chain.

SELECT record\_id\\
FROM observation\_instance oi, observation\_instance coi,
context\_instance ci,
measurement\_instance mi, measurement\_instance cmi \\
WHERE mi.oid=ci.oid AND ci.oid = oi.oid and  cmi.oid=coi.oid AND ci.context\_oid =
coi.oid and oi.otype = 'olabel' and coi.otype $\in$
context\_olabel\_set
AND mi.mvalue=`Picea rubens' and cmi.mvalue='California';\\

>>>>

\section{Introduction}
In this work, we study the query processing over scientific
observation and measurement data using OBOE
model\cite{DBLP:conf/er/BowersMS08}. OBOE model is a conceptual model
used to interpret observation and measurement data.

\subsection{Background}

%Application background
In many scientific domains (e.g., ecology, hydrology, earth science,
geology), people collect observational data. Such data
record the observed value of some real world entity at some specific
place and time. E.g., ecologists studying relationship between the 
growth pattern and the treatments often need to record the tree
heights. The collected data reflect the fact the tree height of a
specific tree is 30.1in on May 1, 2009 and 30.3in on May 1, 2010.

%Characteristics of scientific observational data: non-normalized
Almost all such scientific data do not follow database higher normal
forms. Generally, scientists have their way in interpreting their
data, but they are not ready for any normalization process. 
For example, Table \ref{tb:dataset} is a simplified but typical dataset collected by
a scientist who study the growth pattern of trees. 
Obviously, the ``plt'' of the first two rows is the same place, and
the ``plt'' of the last two rows is the same. 
Here, their area information is redundant. 
In the real application, many columns have redundant information.

%HP: this one has the coordinates, it's troublesome to explain. 
%% \begin{table}[htb]
%% \begin{center}
%% \begin{tabular}{|l|l|l|l|l|}
%% \hline
%% code & no & ht & plt & area\\\hline
%% piru & 1 & 35.8 & $(34^{\circ}8'3N,118^{\circ}14'37W)$ & 4.0\\\hline
%% piru & 2 & 36.2 & $(34^{\circ}8'3N,118^{\circ}14'37W)$ & 4.0\\\hline
%% piru & 1 & 25.7 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0 \\\hline
%% %abba & 1 & 15.6 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0\\\hline
%% capo & 1 & 15.6 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0\\\hline
%% \end{tabular}
%% \end{center}
%% \caption{Dataset}
%% \label{tb:dataset}
%% \end{table}

\begin{table}[htb]
\begin{center}
\begin{tabular}{|l|l|l|l|l|}
\hline
code & no & ht & plt & area\\\hline
piru & 1 & 35.8 & California & 4.0\\\hline
piru & 2 & 36.2 & California & 4.0\\\hline
piru & 1 & 25.7 & Oregon & 3.0 \\\hline
%capo & 1 & 7.8 & Oregon & 3.0\\\hline
Oriental poppy& 1 & 7.8 & Oregon & 3.0\\\hline 
\end{tabular}
\end{center}
\caption{Dataset}
\label{tb:dataset}
\end{table}
%HP: California poppy info
%Oregon iris: http://en.wikipedia.org/wiki/List_of_native_Oregon_plants
%http://en.wikipedia.org/wiki/California_poppy

%Introduce major concepts
In scientific domains where people collect observation and measurement
data, there are several commonly used and widely recognized canonical
concepts(\cite{oboe, om}). 
In this paper, we refer these concepts as OM concepts. 
The canonical concepts include {\em observation}, {\em measurement},
{\em characteristic}, {\em standard}, {\em protocol} or ({\em procedure}). 
For example, {\bf add an example to illustrate these concepts. }

%Introduce annotation, Motivation to use annotation
Generally, every scientific dataset goes into the data repository with
some metadata, e.g., Dublin core metadata\cite{***}, Darwin core metadata\cite{***}. 
However, all these metadata are at the dataset level, they did not
provide enough information for the data {\em content} inside the
dataset. 
To better make use of the data content in the repository, more systems
are embracing the ideas of having metadata on the data content. 
E.g., Some systems \cite{tdar} provides a mechanism to collect the
column/attribute level metadata, some system \cite{semtools} uses
{\em annotation} to add more semantic information to the data
content. In this work, we follow the terminology in \cite{semtools} and
use the term {\em annotation} to distinguish the 
metadata on the data content from those on data objects at a coarser
granularity. 
%How to represent the annotation and the symbols used
We propose tools for scientists to provide annotations to scientific
observational data. 
\from{HP}{Put a screen-dump of annotation.}

%Challenge to query such data
\from{HP}{Queries to such data and challenges to answer such queries.}
When a user formulates a query, it is unrealistic for a him/her to know the 
underlying data structure of the data.  
But they all know the well recognized OM concepts as discussed above. 
So, naturally, when searching such scientific data, people are more
interested in finding datasets related to such OM terminologies as observation
measurement, characteristic, etc. 
Given the dataset in Table \ref{tb:dataset}. 
People may ask the following queries.

\begin{example}\label{eg:query}
Simple and summarization queries: 
\begin{itemize}
\item $Q_1$: Give me the data sets that contain species ``Picea rubens'' observations.
%$Q_2$: Give me the data sets that have measurements on ``area''  characteristics. 
\item $Q_2$: Give me the data sets that contain species ``Picea rubens''
  observations in ``California''. 
\item $Q_3$: Give me the data sets that contain at least five distinct
  ``Picea rubens'' observations.
\item $Q_4$: Give me the data sets that have trees with average ``height''
  than 20.0 in ``California''. 
% Shawn: an example on summarize one measure with respect to another.
% This could take two forms: one where the context forms the groups
% (like by year), and another where the two measurements are within the
% same context (like finding correlations).
\end{itemize}
\end{example}

\from{HP}{Current data integration effort}

\from{HP}{Our effort in querying observational data: introduce OBOE, query}

\from{HP}{Need re-organization. The description in the
  following several sections may be moved to here later.}

@@

\from{HP}{This will come after the real problem definition and the
  solution. The description in the following several sections may be
  moved to here later.}

@@

@@

%Data model: Data, attribute, data value set
When a scientist contributes data into an integrated data repository,
a widely accepted way is to convert each dataset to a data
table\cite{tdar} or treated as a separate object entity\cite{knb}. 
{\bf (HP: is it really widely used? any other system uses this
  way?. Add more citations here.)}
In using this method, the database contains metadata of each
dataset and the definition of the data table (e.g., attribute/column
name, attribute type, etc).

%HP: 
%An alternative way to organize the data is to store them in one big
%data table with three columns data\_file, attribute\_name,
%attribute\_value. The problem with this structure is that the
%attribute\_value are different types.
In our work, we focus on querying this kind of data repository (or
databases). To formalize the scenario, we use $D$ to denote the set of data
tables in the data repository and $d$ to refer to a specific data
table. Each data table $d$ contains metadata about the attribute
definition $Attr_{d}$. Sometimes, we also use {\em column} to refer to an
attribute. 
Given one data table $d$, an attribute $attr_i \in Attr_{d}$ or column index $i$, 
$d[attr_i]$ or $d[i]$ represents the set of values for the attribute
$attr_i$ or for the $i$-th column.
%HP: Implementation detail (no need to put it in the paper): 
%When we put the data to data tables, we add a unique record id for
%each row in the data. This is used to get the unique record id.

\subsection{Annotation}\label{sec:annotation}

%What kinds of annotation information is needed.

We use $A$ to denote the annotation of one dataset. 
Internally, we keep the following information for the
annotation.

For data table $d$ with annotation $A$, the system keeps the
information $AD$ = $\{(d_{id}, A_{id}, d_{meta})\}$.
As we described before, we have four main objects to describe: 
observation type (OT), measurement type (MT), and
context type (CT), and the mapping (Map) from the measurement type to
resource attributes.  
So, the annotation contains the following information.

\begin{itemize}
\item $A.OT$ = $\{(\underline{A_{id},ot_{id}}, et, isDistinct)\}$ to
  describe an observation type, which denotes on which
  entity type (object in the real world) the observation is made. 
Very often, more than one observation can be made on one entity. Thus,{\em
  isDistinct} is used to denote whether the same value of key
meaasurements of an obbservation types can uniquely identify one
observation or not. 
\item $A.MT$ = $\{(\underline{A_{id}, mt_{id}}, ot_{id},isKey,
  Cha, \cdots)\}$. Generally, $MT$ contains information about
  characteristic ($Cha$), Standard, Protocol and Precisio. We do not include them
  here to make the description clearer. Here, {\em isKey} is used to denote
  whether one measurement type is the key measurement for the
  observation type $OT_{id}$ or not. 
\item $A.CT$ = $\{(\underline{A_{id}, ot_{id}, cot_{id},
  Rel}, isIdentify)\}$ where $cot$ represents the context observation
  type. 
\item $A.Map$ = $\{(\underline{A_{id},mt_{id}, attr_i, Cond},
  Val\}$
\end{itemize}

We use the following example to illustrate the concepts in the
annotation.

\begin{example}
$A.OT$ = \{($A_1, OT_1$, tree, false),($A_1, OT_2$, GeoSpot, true) \}
  denotes that annotation $A_1$ are for two observation types. One
  type $OT_1$ is on real world entity ``tree'' and the values of its key measurements
  do not uniquely identify a distinct observation. The other type
  $OT_2$ is for real world entity ``GeoSpot'' (i.e., geospatial
  location) and its key measurement types can identify it's unique
  observations. \from{HP}{Introduce key measurements before this.}

$A.MT$ =\{($A_1, MT_1, OT_1$, Species, true), 
($A_1, MT_2, OT_1$, SpecNo, true), 
($A_1, MT_3, OT_1$, Height, false),
($A_1, MT_4, OT_2$, Plot\_State, true), 
($A_1, MT_5, OT_2$, Plot\_Area, false)\} represents that $OT_1$
(for ``tree'') has three measurement types ($MT_1, MT_2$, $MT_3$) for characteristics
``Species'', ``SpecNo'', and ``Height''. The fist two measurement
  types together form the key measurement type for $OT_1$. 
$OT_2$ (for ``GeoSpot'') has one key measurement type $MT_4$ for
  characteristic ``Plot\_State'' and another measurement type $MT_5$
  for characteristic ``Plot\_Area''.

$A.CT$ = \{($A_1$,$OT_1$, $OT_2$, within, true)\}. It shows that
observation of a {\em tree} is within the context of a geo-spatial
location. And this context is used to identify the uniqueness of the
observation. E.g., a tree with species name ``piru'' and species no
$1$ in California
%$(34^{\circ}8'3N,118^{\circ}14'37W)$ (Los Angeles in California) 
is different from a tree with 
the same species name ``piru'' at Oregon. 
%location $36^{\circ}10'30N,115^{\circ}8'11W$ (Las Vegas in
%California).

%HP coordinates material: 
%http://www.traveljournals.net/explore/united_states/map/u1662328/los_angeles.html
%http://en.wikipedia.org/wiki/Las_Vegas,_Nevada

$A.Map$ = \{($A_1$, $MT_1$, ``code'', ``eq `piru''', ``Picea rubens''),
%($A_1$, $MT_1$, ``code'', ``eq `abba''', ``Abies balsamea''),
%($A_1$, $MT_1$, ``code'', ``eq `capo''', ``California poppy''),
($A_1$, $MT_2$,``no'', null,), 
($A_1$, $MT_3$, ``plt'', null,), 
($A_1$, $MT_4$,``area'', null,\}. 
The first mapping rules maps ``code''
attribute to the measurement type $MT_1$ (for ``Species''
characteristic) and change the value to ``Picea rubens'' if the code
is ``piru''. The meaning of the other mapping rules are very
obvious here, so we skip the details. 
\end{example}

\subsection{Observation and measurement (OM) query}
Above we defined the data model and the annotation, in this
sub-section, we define the queries that scientists are interested in
We denote the queries on the observational data as {\em Observation
and measurement (OM) query}.

In using the observational data, people are generally interested in
the following concepts: {\em observation, measurement,characteristic or standard}.

\begin{definition}[Basic OM query]\label{def:basic_omq}
A basic observation and measurement query is defined as $Q$::={\tt
  concept:cond}. 
Here, {\tt concept} is the main term in an observation model. 
{\tt cond} is in the form of $f([distinct]~attr)~op~value$ where $op$ is
basic operator from $\{=, \neq, >, <\}$ and $f$ is an
aggregation function from $\{sum, avg, count, min, max\}$. 
\end{definition}

\begin{definition}\label{def:qresult}
The result of a query $Q$ is a set of data objects (e.g., data tables) 
$\{d|d\in D \wedge d~satisfies~{\tt cond}\}$. 
For each of such result $d$, we use $d~s.t.~Q$ to denote that
data object $d$ satisfies the query. 
\end{definition}

Based on the different {\tt cond} definition, ``$d~satisfies~{\tt cond}$''
is translated into different formulas. 
E.g., if {\tt cond} is defined that the value of attribute ``area'' need to be
smaller than $3.0$, then, ``$d~satisfies~{\tt cond}$'' is translated to
``$dv<3.0|dv\in d[area]$''.

Definition \ref{def:basic_omq} can formulate queries on the
one specific observation and measurement. 
In real application, we need to consider the context relationship. 
So, we generalize the basic OM query to contextualized OM query as
follows.

@@

For example, the queries in Example \ref{eg:query} can be formalized
as the following formal OM queries.

%\begin{example}
\begin{itemize}
\item $Observation: \langle Species='Picea~rubens'\rangle$
\item $Observation_1: \langle Species='Picea~rubens'\rangle$\\
  $IN(Observation_1, Observation_2$) \\
$Observation_2: \langle Meas_{unknown}=`California'\rangle$
\item $Observation: \langle Species='Picea~rubens' $\\
$\wedge
  count(distinct~Species)\geq 5\rangle$
\item $??$
\end{itemize}

%\end{example}

%\from{HP}{Add experiement: show the error rate (e.g., precision, recall, (or
%  directly $F_1$) of this method.? not needed?}

@@

In this section, we discuss several methods that we can use to
answer OM queries over scientific observational data with
annotations.

%\subsection{Naive method}\label{sec:naive}

A very direct way search such data is to extract all the content in
the data cells and index them, this way, the system may help answer
very simple queries with the help of some thesaurus. 
For example, for $Q_1$ in Example \ref{eg:query}, as long as we can
translate the abbreviation ``piru'' in the dataset to ``Picea rubens''
based on some predefined thesaurus. We can find the results. 
Even for this kind of simple query, however, if we want to find
observation of ``California poppy'', an approximate match may return
the dataset in Table \ref{tb:dataset} as a result because it contains
both {\em California} and {\em Poppy}.
If such compound word search can be alleviated using some existing
techniques \cite{***}, %citation on compound word querying. 
it is almost impossible to directly use such IR-style system to answer a query
in the form of $Q_3$ and $Q_4$. 
I.e., answer queries with context.

This naive method cannot answer the desired query. It is because it
fails to catch the semantics in the data. 
In what follows, we would use the annotation during our search to get
more semantic support.

As we analyzed in Section \ref{sec:relatedwork}, there are two extrems
in querying integrated data. 
One is to rewrite the original query to a series of new queries over
the data; the other is to materialize the data to a consistent data
model and then answer the query over the materialized database. 
We what follows we work from these two directions to leverage the
semantic information in answering such queries.

@@

As described in Section \ref{sec:dataquery}, 
an OM query is represented using 
OM concept terminologies such as observation
measurement, characteristic, etc. 
To answer such query from the original data model (i.e., dataset
metadata and data tables), we need to {\em rewrite} the quey over the
real data table. 
In what follows, we first sketch the re-writing process. Then, we
detailed the procedure in using the data model and
structures. Finally, we include the process for the more complicated
cases (e.g., with distinct, aggregate).

@@

The first step is to map the OM query to the real data structure. 
We can utilize the annotation structure $A.Map$ which keeps the correspondences between
measurement type and table attributes. 
So, when the given concept is {\em measurement}, we can directly get
its related attribute names and annotation id $A_{id}$. 
In case that the given OM concept is a non-measurement concept, we can use
other Annotation structures $A.*$ to figure out the measurementt types. 
E.g., if the OM concept is Characteristic ``Species'', we can use
$A.MeasType$ to get the measurement type id.

As mentioned in Section \ref{sec:annotation}, when some annotations are done over data tables, the system keeps the
information $\langle A_{id}, DTableId\rangle$. 
With the $A_{id}$, we can ge the data table information from the
metadata  $\langle A_{id}, DTableId\rangle$. 
Then, we get the relevant data table information and the needed attribute
information.

@@

The basic idea is to materialize the data into some centralized
concept tables with instances of entities, observations, measurements
and the contexts between observations. 
We call such concept tables core OBOE tables, and denote them as
$OBOE.*$. 
In partibular, we have the following concept instance tables. 
\begin{itemize}
\item $OBOE.EI$ = $\{(\underline{ei_{id}}, et_{id})\}$ for all the entity instances;
\item $OBOE.OI$ = $\{(\underline{oi_{id}}, ot_{id}, ei_{id})\}$ keeps
  all the observation instances;
\item $OBOE.MI$ = $\{(\underline{mi_{id}}, oi_{id}, mt_{id},
  mVal)\}$ for all the measurement instances;
\item $OBOE.CI =\{(\underline{oi_{id},coi_{id},ct_{id}})\}$ for
  all the context instances;
\end{itemize}

@@

\begin{example} \label{eg:materialize_db_q1}
Example \ref{eg:query_reqerite_q1} shows how to deal with $Q_1$ using
the query rewriting method. 
Using the  materialized database, the first step is the same.
After we know the data file, then we search the OBOE.MI with the
condition that {\em mVal=``Picea rubens''}.
\end{example}

{\tt 
SELECT distinct $AD.d_{id}$\\
FROM MI, MT, AD\\ 
WHERE $MT.Cha='Species'$ \\
AND $MI.mVal = `Picea rubens'$\\
AND $MI.mt_{id}=MT.mt_{id}$ AND $MT.A_{id}=AD.A_{id}$; %\#Join the MI and MT tables
}

In this example, the computation cost is the selection cost of the
measurement instance table. 
It should be faster than the query rewriting.

If the query has some restriction on the aggregate result, we can
easily add the GROUP BY and HAVING clause to deal with it. 
We take $Q_3$ as an example, which asks for data sets that contain at
least five distinct ``Picea rubens'' observations.
The SQL query can be written as: \\
{\tt 
SELECT distinct $d_{id}$\\
FROM MI, MT\\ 
WHERE $MT.Cha='Species'$ \\
AND $MI.mVal = `Picea rubens'$\\
AND $MI.mt_{id}=MT.mt_{id}$\\
GROUP BY $d_{id},oi_{id}$\\
HAVING COUNT(*)$>$5; %\#Join the MI and MT tables
}

Using the query rewriting, 
First, we find the data tables that contain the needed concepts. 
In each table, we need to figure out the distinct observations. 
From $m_i$ $obstype_i$, all the other measurements that are of
$obstype_i$. 
Group by the key measurements of $obstype_i$ if distinct yes.
If this observation type does not have any context, then the key
measurements are all the measurements directly defined under this
observatioin type. Otherwise, i.e., this observation type has context,
then the key measurements are all the measurements in the context
chain. 
%\end{example}

When we are given contextualized query, e.g., $Q_2$. 
Give me the datasets that contain species ``Picea rubens'' observations in ``California''.

Species = ``Picea rubens'' and observation has context {\em in}
another observation which has measurement of value ``California''.

%Using materializedDB, we can answer the query this way
%First, from the Characteristic ``Species'', we can get  --> mlabel, observation type

%select mtype, olabel from measurement\_type  \\
%where chracteristic = 'Species';  \\

{\tt 
SELECT record\_id\\
FROM CI as ci, MI as mi, MI as cmi,MT as mt, MT as cmt, \\
WHERE $mi.oi_{id}=ci.oi_{id}$ AND $cmi.oi_{id}=ci.coi_{id}$ \\
AND $ci.coi_{id} = coi.oi_{id}$ AND oi.otype = 'olabel' \\
AND mi.mvalue=`Picea rubens' AND cmi.mvalue='California'\\
AND MT.mt_{id}=mi.mt_{id} AND MT.mLabel = 'Species'\\
AND CMT.mt_{id}=cmi.mt_{id} AND cmt.mLabel=`State'\\
AND coi.otype IN (\\
\verb|    |SELECT context\_olabel\\
\verb|    |FROM context\_relationship\\
\verb|    |WHERE olabel = 'olabel');
}
The sub-query finds the set containing all the observation type
labels in the context chain.
